A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices

### 2013-10-19 tex/LanczosProcedures.tex
minor formatting mods
miscellaneous fixes

\chapter{Lanczos Procedures}

The following will review the usual procedure for generating an orthonormal basis for
$\sK(\M, \psi_1, p)$, 
and conclude by showing that, for our problem, we can find the matrix 
$\Q^t\M\Q$
without actually carrying out this procedure. Note that this conclusion is essential since the procedure
requires matrix vector multiplication---an operation we have assumed impossible for large enough $d$.

@@

Note that Equation~(\ref{eq:4.3}) results from the equation $\M\Q = \Q\T$. The latter equation holds
since, starting with $d$ vectors in our subspace, $\Q$ is a $d \times d$ matrix
whose columns form an orthonormal basis for all of $\R^d$, which is clearly an
invariant subspace. The Lanczos algorithm above, however, 
%
%
% ------------ 25.txt ---------------------------------------------------------
%
%
proceeds for only $p$ steps, producing a $d \times p$ matrix $\Q$, whose columns
form an orthonormal basis for the \emph{approximate} invariant subspace 
$\sK(\M, \psi, p)$, and a $p \times p$ matrix $\T, = \Q^t\M\Q$. In that case,
$\|\M\Q - \Q\T_p\|_2 = \|\E\|_2$ is nonzero and gives the error bound described
in Theorem~\ref{thm:3.2.1}. Now, writing the full $d \times d$ matrix $\T$
of~(\ref{eq:4.2}) as
\begin{equation}
  \label{eq:4.4}
\T = 
\left( \begin{array}{c|c}
\T_p & \T_{pu}^t \\
 \hline
\T_{pu} & \T_{u}
\end{array}
\right)
=
\left( \begin{array}{cccc|cccc}
\alpha_1 & \beta_1 &      &                 & & && \\
\beta_1 & \ddots  & \ddots&                 & & && \\
        & \ddots  & \ddots  & \beta_{p-1}    &          & && \\
        &         & \beta_{p-1}  & \alpha_{p} & \beta_{p}& && \\
\hline
        &         &             & \beta_{p}  & \alpha_{p+1} & \beta_{p+1} && \\
        &         &             &           & \beta_{p+1}  & \ddots & \ddots &\\
        &         &             &           &             & \ddots & \ddots &\beta_{d-1} \\
        &         &             &           &             &        & \beta_{d-1} &\alpha_{d}
\end{array}
\right)
\end{equation}
allows us to describe the error bound $\|\E\|_2$ in terms of the submatrix
$\T_{pu}$, and hence in terms of $\beta_p$.
\begin{theorem}
%Theorem 4.1.1
\label{thm:4.1.1}
If $\T_p$ and $\T_{pu}$ are the matrices appearing in~(\ref{eq:4.4}), and if the $p$
columns of $\Q$ are computed by the 
Lanczos algorithm, then there exist $\mu_1, \dots, \mu_p \in \lambda(\T_p)$ and 
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$
such that
\[
|\mu_k - \lambda_k| \leq \|\T_{pu}\|_2 = \beta_p, \quad \text{ for $k=1,\dots,p$.}
\]
\end{theorem}
For a proof, see~\cite[Page 365]{Demmel:1997}.
%
%
% --------------- 26.txt -------------------------
%
%
\section{A Lanczos Procedure for Markov Chains}
Notice that the Lanczos algorithm requires the computation of $\M q_j$. Throughout this
paper we have assumed that $\M \in \R^{d\times d}$, and that $d$ is so large as
to make the matrix vector multiplication $\M q_j$ impossible. Now, the Lanczos
algorithm constructs an orthonormal basis for the Krylov subspace. For our
purposes, this construction is not essential. What is essential is that we find
a way to generate the Lanczos coeficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
whence the matrix $\T$, without performing the operation $\M q_j$ required by the
Lanczos algorithm. We now address this problem.

@@

Before generalizing this for the $j$th Lanczos coefficient, we consider one more coeﬂicient
in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
$q_2 = (\M q_1 — \alpha_1 q_1)/\beta_1$. 
From this we can derive an expression for $\alpha_2$ which, while not as neat as those
%
%
% ----------------- 28.txt ----------------------------------------------------------
%
%
for $\alpha_1$ and $\beta_1$, is nonetheless tractable.
\begin{align}
\label{eq:4.8}
\alpha_2 &= q_2^t \M q_2 \nonumber\\
&= \frac{1}{\beta_1^2}(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
\end{align}
Equation~(\ref{eq:4.8}) follows from Equation~(\ref{eq:4.7}), which can be
written as $q_1^t\M^2 q_1 = \alpha_1^2 + \beta_1^2$.  
The preceding paragraphs define all but one of the quantities in the last
exression of~(\ref{eq:4.8}). The one remaining
is $q_1^t\M^3 q_1$, which can also be expressed in terms of covariances of $\phi$,
and the analysis is almost identical to that for $q_1^t\M^2q_1$:
\begin{align*}
q_1^t\M^3 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^3(\phi- \bE_\pi \phi)}
{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\
&= \frac{\bC_\pi (\phi(n), \phi(n + 3))}{\Var_\pi(\phi(n))}.
\end{align*}
To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
relation between $(\alpha_{j+1}, \beta_j)$ and $(\alpha_{j}, \beta_{j-1})$. 
As we will see, such a relation involves higher order
Lanczos vectors $q_j$---vectors we do not wish to compute. Recall that the only ``vector'' we can
handle is our initial observable. Therefore, we must also establish a recurrence
among the $q_j$'s, thus making them available inductively from our observable vector.

@@

The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
these vectors will not be readily available. However, another recursion---this time relating forms
involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
expressions accessible by induction on our initial observable. The general forms
of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
begin with the simpler of the two:
\begin{align*}
  q_{j+1}^tM^kq_j &= 
 \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
 \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
\end{align*}
The second is,
\begin{align}
\label{eq:4.12}
  q_{j+1}^tM^kq_{j+1} 
&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} 
    - 2\alpha_jq_j^t\M^{k+1} q_j
    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
\end{align}
The point of all this is that each pair of coeﬂicients $(\alpha_{j+1}, \beta_j)$
can be expressed as a function of the covariance between 
$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
defining these recursions in a computer program is trivial, and we use this fact
in Section~\label{sec:computations} % Sec 5.4
when applying the foregoing to a particular problem.
%
%
% ----------------- 30.txt ------------------------------------------------------
%
%

>>>>

\chapter{Lanczos Procedures}
\label{cha:lanczos-procedures}
The following will review the usual procedure for generating an orthonormal basis for
$\sK(\M, \psi_1, p)$, 
and conclude by showing that, for our problem, we can find the matrix 
$\Q^t\M\Q$
without actually carrying out this procedure. Note that this conclusion is essential since the procedure
requires matrix vector multiplication---an operation we have assumed impossible for large enough $d$.

@@

Note that Equation~(\ref{eq:4.3}) results from the equation $\M\Q = \Q\T$. The latter equation holds
since, starting with $d$ vectors in our subspace, $\Q$ is a $d \times d$ matrix
whose columns form an orthonormal basis for all of $\R^d$, which is clearly an
invariant subspace. The Lanczos algorithm above, however, 
%
%
% ------------ 25.txt ---------------------------------------------------------
%
%
proceeds for only $p$ steps, producing a $d \times p$ matrix $\Q$, whose columns
form an orthonormal basis for the \emph{approximate} invariant subspace 
$\sK(\M, \psi, p)$, and a $p \times p$ matrix $\T, = \Q^t\M\Q$. In that case,
$\|\M\Q - \Q\T_p\|_2 = \|\E\|_2$ is nonzero and gives the error bound described
in Theorem~\ref{thm:3.2.1}. Now, writing the full $d \times d$ matrix $\T$
of~(\ref{eq:4.2}) as
\begin{equation}
  \label{eq:4.4}
\T = 
\left( \begin{array}{c|c}
\T_p & \T_{pu}^t \\
 \hline
\T_{pu} & \T_{u}
\end{array}
\right)
=
\left( \begin{array}{cccc|cccc}
\alpha_1 & \beta_1 &      &                 & & && \\
\beta_1 & \ddots  & \ddots&                 & & && \\
        & \ddots  & \ddots  & \beta_{p-1}    &          & && \\
        &         & \beta_{p-1}  & \alpha_{p} & \beta_{p}& && \\
\hline
        &         &             & \beta_{p}  & \alpha_{p+1} & \beta_{p+1} && \\
        &         &             &           & \beta_{p+1}  & \ddots & \ddots &\\
        &         &             &           &             & \ddots & \ddots &\beta_{d-1} \\
        &         &             &           &             &        & \beta_{d-1} &\alpha_{d}
\end{array}
\right)
\end{equation}
allows us to describe the error bound $\|\E\|_2$ in terms of the submatrix
$\T_{pu}$, and hence in terms of $\beta_p$.
\begin{theorem}
%Theorem 4.1.1
\label{thm:4.1.1}
If $\T_p$ and $\T_{pu}$ are the matrices appearing in~(\ref{eq:4.4}), and if the $p$
columns of $\Q$ are computed by the 
Lanczos algorithm, then there exist $\mu_1, \dots, \mu_p \in \lambda(\T_p)$ and 
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$
such that
\[
|\mu_k - \lambda_k| \leq \|\T_{pu}\|_2 = \beta_p, \quad \text{ for $k=1,\dots,p$.}
\]
\end{theorem}
For a proof, see~\cite[Page 365]{Demmel:1997}.
%
%
% --------------- 26.txt -------------------------
%
%
\section{A Lanczos Procedure for Markov Chains}
% Sec 4.2
\label{sec:lancz-proc-mark}
Notice that the Lanczos algorithm requires the computation of $\M q_j$. Throughout this
paper we have assumed that $\M \in \R^{d\times d}$, and that $d$ is so large as
to make the matrix vector multiplication $\M q_j$ impossible. Now, the Lanczos
algorithm constructs an orthonormal basis for the Krylov subspace. For our
purposes, this construction is not essential. What is essential is that we find
a way to generate the Lanczos coeficients $\alpha_i$, $\beta_j$, $(j = 1,\dots, p)$, and
whence the matrix $\T$, without performing the operation $\M q_j$ required by the
Lanczos algorithm. We now address this problem.

@@

Before generalizing this for the $j$th Lanczos coefficient, we consider one more coeﬂicient
in detail. Recall that the second basis element $q_2$ in the Lanczos algorithm is given by 
$q_2 = (\M q_1 — \alpha_1 q_1)/\beta_1$. 
From this we can derive an expression for $\alpha_2$ which, while not as neat as those
%
%
% ----------------- 28.txt ----------------------------------------------------------
%
%
for $\alpha_1$ and $\beta_1$, is nonetheless tractable.
\begin{align}
\label{eq:4.8}
\alpha_2 &= q_2^t \M q_2 \nonumber\\
&= \frac{1}{\beta_1^2}(\M q_1 - \alpha_1 q_1)^t\M (\M q_1 - \alpha_1 q_1)\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 q_1^t\M^2 q_1 + \alpha_1^2q_1^t\M q_1 )\nonumber\\
&= \frac{1}{\beta_1^2}(q_1^t\M^3 q_1 - 2\alpha_1 \beta_1^2 + \alpha_1^3 )
%&= \bigl[q_1^t\M^2 q_1 - \alpha^2_1\bigr]^{1/2}.
\end{align}
Equation~(\ref{eq:4.8}) follows from Equation~(\ref{eq:4.7}), which can be
written as $q_1^t\M^2 q_1 = \alpha_1^2 + \beta_1^2$.  
The preceding paragraphs define all but one of the quantities in the last
exression of~(\ref{eq:4.8}). The one remaining
is $q_1^t\M^3 q_1$, which can also be expressed in terms of covariances of $\phi$,
and the analysis is almost identical to that for $q_1^t\M^2q_1$:
\begin{align*}
q_1^t\M^3 q_1  &= \frac{(\phi - \bE_\pi\phi)^t\Pi \P^3(\phi- \bE_\pi \phi)}
{(\phi - \bE_\pi\phi)^t\Pi (\phi- \bE_\pi \phi)} \\
&= \frac{\bC_\pi (\phi(n), \phi(n + 3))}{\Var_\pi(\phi(n))}.
\end{align*}
To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
relation between $(\alpha_{j+1}, \beta_j)$ and $(\alpha_{j}, \beta_{j-1})$. 
As we will see, such a relation involves higher order
Lanczos vectors $q_j$---vectors we do not wish to compute. Recall that the only ``vector'' we can
handle is our initial observable. Therefore, we must also establish a recurrence
among the $q_j$'s, thus making them available inductively from our observable vector.

@@

The expressions above involve the higher order Lanczos vectors $q_j$. In the present context,
these vectors will not be readily available. However, another recursion---this time relating forms
involving $q_{j+1}$ to forms involving $q_j$ and $q_{j-1} $---will make these
expressions accessible by induction on our initial observable. The general forms
of interest are $q_{j+1}^t\M^k q_{j+1}$ and $q^t_{j+1}\M^k q_j$. Again, we 
begin with the simpler of the two:
\begin{align*}
  q_{j+1}^tM^kq_j &= 
 \frac{1}{\beta_j}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k q_j\\
 \frac{1}{\beta_j}(q_j^t\M^{k+1} q_j - \alpha_j q_j^t\M^{k} q_j -\beta_{j-1} q_j^t\M^k).
\end{align*}
The second is,
\begin{align}
\label{eq:4.12}
  q_{j+1}^tM^kq_{j+1} 
&=  \frac{1}{\beta_j^2}(\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})^t\M^k 
  (\M q_j - \alpha_j q_j - \beta_{j-1} q_{j-1})\nonumber\\
&=  \frac{1}{\beta_j^2}(q_j^t\M^{k+2} q_j + \alpha^2_jq_j^t\M^{k} q_j 
  + \beta^2_{j-1} q_{j-1}^t\M^k q_{j-1} - 2\alpha_jq_j^t\M^{k+1} q_j \nonumber\\
& \qquad    
    + 2\alpha_j \beta_{j-1}q_j^t\M^k q_{j-1}
    - 2\beta_{j-1}q_j^t \M^{k+1}q_{j-1}).
\end{align}
The point of all this is that each pair of coeﬂicients $(\alpha_{j+1}, \beta_j)$
can be expressed as a function of the covariance between 
$\phi(n)$ and $\phi(n+ m)$, for $m = 0, \dots, 2j+1$. Since the expressions
are based on four recurrence relations, they become quite complicated as $j$ gets large. However,
defining these recursions in a computer program is trivial, and we use this fact
in Section~\label{sec:computations} % Sec 5.4
when applying the foregoing to a particular problem.
%
%
% ----------------- 30.txt ------------------------------------------------------
%
%

### 2013-10-19 tex/intro.tex
miscellaneous fixes

The rate at which a Markov chain converges to a given probability distribution has long
been an active area of research. This is not surprising considering this problem’s relevance to the ar-
eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
algorithms provide important examples. These algorithms come in handy when we encounter a
complicated probability distribution from which we want to draw random samples. In statistical
mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
Geman and Geman

To implement the MCMC algorithm, we invent a Markov chain that converges to the
desired distribution (this is often accomplished using the Metropolis algorithm
described in Chapter 5). Realizations of the chain will eventually represent
samples from this distribution. Sometimes ``eventually'' -- meaning all but
finitely many terms of the chain -- is just not enough. We need more practical
results. In particular, we want to know how many terms of the chain should be 
discarded before we are sampling from a distribution that is close (in total variation distance) to
% 03.txt
the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.

Often the Markov chains encountered in this context satisfy a condition known in the
physics literature as detailed balance. Probabilists call chains with this property reversible. This
simply means that the chain has the same probability law whether time moves
forward or 
backward.\footnote{This is not a precise definition. In particular the chain must
  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}

There are a number of different methods in common use for bounding convergence rates of
Markov chains, and a good review of these methods with many references can be found in  More
recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
probability matrix, and thus require good approximations to such eigenvalues. In many applications,
however, the transition probability matrix is so large that it becomes impossible to store even a
single vector of the matrix in conventional computer memory. These so called out-of-core problems
are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
without modification. This paper develops 
such a modification for the Markov chain eigenvalue problem. In particular it develops a method
for approximating the first few eigenvalues of a transition probability matrix when we know the
general structure of the underlying Markov chain. The method does not require storage of large
matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
analysis of the simulation.

Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
% 04.txt
space of the Markov process. This section and Chapter 3 develop the context in which we formulate
the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
Krylov subspace and explain why this represents our best approximation to a subspace containing
extremal eigenvectors of the transition probability matrix.\footnote{or, more
  precisely, a similarity transformation of this matrix.} The first section of
Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
for the Krylov subspace. As it stands, this algorithm is useless for an
out-of-core problem such as ours since, by definition of such problems, 
it requires too much data movement; all the computing time is spent swapping data between slow
and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
It is here that we experiment with the procedure described in Section 4.2 and approximate the
extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
the paper.
% 05.txt

>>>>

The rate at which a Markov chain converges to a given probability distribution
has long been an active area of research. This is not surprising considering
this problem's relevance to the areas of statistics, statistical mechanics, and
computer science. Markov Chain Monte Carlo (MCMC) algorithms provide important
examples. These algorithms come in handy when we encounter a complicated
probability distribution from which we want to draw random samples. In
statistical mechanics, we might wish to estimate the phase average of a function
on the state space. Goodman and Sokal~\cite{GoodmanSokal:1989} examine Monte
Carlo methods in this context. Examples from statistics occur in the Bayesian
paradigm when we are forced to simulate an unwieldy posterior distribution (see,
e.g., Geman and Geman~\cite{Geman:1984}.

To implement the MCMC algorithm, we invent a Markov chain that converges to the
desired distribution (this is often accomplished using the \emph{Metropolis algorithm}
described in Chapter~\ref{cha:an-appl-conv}). Realizations of the chain will
eventually represent samples from this distribution. Sometimes
``eventually''---meaning all but finitely many terms of the chain---is just not 
enough. We need more practical results. In particular, we want to know how many
terms of the chain should be discarded before we are sampling from a
distribution that is close (in total variation distance) to 
%
%
% ------------- 03.txt -----------------------------------------------------
%
%
the distribution of interest. This is the purpose of bounding convergence rates
for Markov chains.

Often the Markov chains encountered in this context satisfy a condition known in
the physics literature as \emph{detailed balance}. Probabilists call chains with
this property \emph{reversible}. This simply means that the chain has the same
probability law whether time moves forward or backward.\footnote{This is not a
precise definition. In particular the chain must have started from its
stationary distribution. Full rigor is postponed until
Section~\ref{sec:general-theory}.} 
In this paper, we consider the rate at which such chains converge to
a \emph{stationary distribution}. This and other italicized terms are
defined in Section~\ref{sec:general-theory}.

There are a number of different methods in common use for bounding convergence
rates of Markov chains, and a good review of these methods with many references
can be found in~\cite{Rosenthal:1995}.  More recently developed methods
employing logarithmic Sobolev inequalities are reviewed in~\cite{Diaconis:1996}.  
Most of the bounds in common use involve the subdominant eigenvalue of the
Markov chain's transition probability matrix, and thus require good
approximations to such eigenvalues. In many applications, however, the
transition probability matrix is so large that it becomes impossible to store
even a single vector of the matrix in conventional computer memory. These so
called \emph{out-of-core} problems are not amenable to traditional eigenvalue
algorithms without modification.\footnote{By ``traditional 
  eigenvalue algorithms'' we mean those found, for example, in Golub and Van
  Loan~\cite{Golub:1996}. See also the book by Demmel\cite{Demmel:1997} for more 
  recent treatment.}
In this paper we develop such a modification for the Markov chain eigenvalue
problem. In particular we develop a method for approximating the first few
eigenvalues of a transition probability matrix when we know the general
structure of the underlying Markov chain. The method does not require storage
of large matrices or vectors. Instead we need only simulate the Markov chain,
and conduct a statistical analysis of the simulation.

Here is a brief summary of the paper. Section~\ref{sec:general-theory} contains
a review of the relevant Markov chain theory. Readers who already know the
basics of asymptotic theory of Markov chains might wish to skim 
Section\ref{sec:general-theory} if only to familiarize themselves with our
notation. Section~\ref{sec:funct-state-space} describes functions on the state 
% 04.txt
space of the Markov process. This section and Chapter~\ref{cha:invar-appr-invar}
develop the context in which we formulate the new ideas of the paper. In the
last section of Chapter~\ref{cha:invar-appr-invar},
Section~\ref{sec:krylov-subspace}, we describe the \emph{Krylov subspace} and
explain why this represents our best approximation to a subspace containing 
extremal eigenvectors of the transition probability matrix (more
  precisely, a similarity transformation of this matrix). The first section of
Chapter~\ref{cha:lanczos-procedures} describes the \emph{Lanczos algorithm} for
generating an orthonormal basis for the Krylov subspace. As it stands, this
algorithm is useless for an out-of-core problem since, by
definition of such problems, it requires too much data movement; all the
computing time is spent swapping data between slow and fast memory (disk to ram
to cache and back). We develop alternatives to the Lanczos algorithm and
demonstrate that the \emph{Lanczos coefficients} of the algorithm can be
obtained by simulations of the Markov chain, and this allows us to avoid
the standard algorithm altogether. Following this is a chapter describing the
Metropolis algorithm used to produce a reversible stochastic matrix. It is here
that we experiment with the procedure described in
Section~\ref{sec:lancz-proc-mark} and approximate the extremal eigenvalues of
the matrix, without storing any of its vectors. Finally,
Chapter~\ref{cha:conclusion} concludes the paper. 
% 05.txt
